---
title: "NBS component analysis"
author: "Lyron Winderbaum"
date: "2021-03-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

\usepackage{bm}

```{r setup}
# # Cleanup and Global Settings
# rm(list = ls())
# if (!is.null(sessionInfo()$otherPkgs)) {
#   invisible(lapply(paste0('package:', names(sessionInfo()$otherPkgs)), 
#                    detach, character.only=TRUE, unload=TRUE))
# }
# graphics.off()
# options(stringsAsFactors = FALSE)

library(tidyverse)
```


## Conceptual Framework

\newcommand{\jb}{j^*}
\newcommand{\kb}{k^*}

First, some notation. Let: 

  * $p_{ij}$ denote an indicator function for the presence absence values 
  with value $1$ for presence and $0$ for absence of gene $j$ in line $i$, 
  * $y_i$ denote the yield of line $i$, 
  * $n$ denote the number of lines and $d$ denote the number of genes,
  * $n_j = \sum_ip_{ij}$ denote the number of lines with presence of gene $j$,
  * $n_{\jb} = \sum_i(1 - p_{ij})$ denote the number of lines with absence of 
  gene $j$,
  * $\bar{y} = \frac{1}{n}\sum_iy_i$ denote the mean yield, 
  * $\bar{y}_j = \frac{1}{n_j}\sum_ip_{ij}y_i$ denote the mean yield 
  of lines with presence of gene $j$, and
  * $\bar{y}_{\jb} = \frac{1}{n_{\jb}}\sum_i(1 - p_{ij})y_i$ denote the 
  mean yield of lines with absence of gene $j$.

Note that 
$\bar{y} = \frac{n_j}{n} \bar{y}_j + \frac{n_{\jb}}{n} \bar{y}_{\jb}$
and that $n = n_j + n_{\jb}$ for all $j$.

We can created coded yield data which here we will explore in more detail: 

$$ x_{ij} = p_{ij}\bar{y}_j + (1 - p_{ij})\bar{y}_{\jb} $$

Note that
\begin{align*}
\frac{1}{n}\sum_ix_{ij} 
  &= \frac{1}{n}\sum_i(p_{ij}\bar{y}_j + (1 - p_{ij})\bar{y}_{\jb}) \\
  &= \frac{\bar{y}_j}{n}\sum_ip_{ij} + \frac{\bar{y}_{\jb}}{n}\sum_i(1 - p_{ij}) \\
  &= \frac{n_j}{n} \bar{y}_j + \frac{n_{\jb}}{n} \bar{y}_{\jb} \\
  &= \bar{y}
\end{align*}
for all $j$.

Let $X_j$ denote the column vector with entries $[x_{ij}]_i$. If we consider the 
yield data to be pre-centered to have $\bar{y} = 0$ then 
the $X_j$ will also be centered by construction for all $j$. One important 
consequence of this centering is that 
$\frac{n_j}{n} \bar{y}_j + \frac{n_{\jb}}{n} \bar{y}_{\jb} = 0$ and so 
$\bar{y}_{\jb} = -\frac{n_j}{n_{\jb}} \bar{y}_j$. We will consider the $y_i$ 
to be centered from here on. The variance of the $X_j$ would then be 
proportional to
\begin{align*}
X_j^TX_j &= \sum_ix_{ij}^2 \\
         &= \sum_i(p_{ij}\bar{y}_j + (1 - p_{ij})\bar{y}_{\jb})^2 \\
         &= \sum_i\left(p_{ij}^2\bar{y}_j^2 + 2p_{ij}(1 - p_{ij})\bar{y}_j\bar{y}_{\jb} + (1 - p_{ij})^2\bar{y}_{\jb}^2\right) \\
         &= \sum_ip_{ij}\bar{y}_j^2 + \sum_i(1 - p_{ij})\bar{y}_{\jb}^2 \\
         &= n_j\bar{y}_j^2 + n_{\jb}\bar{y}_{\jb}^2 \\
         &= n_j\bar{y}_j^2 + n_{\jb}(-\frac{n_j}{n_{\jb}} \bar{y}_j)^2 \\
         &= n_j\bar{y}_j^2 (1 + \frac{n_j}{n_{\jb}}) \\
         &= \frac{n_j}{n_{\jb}}\bar{y}_j^2 (n_{\jb} + n_j) \\
         &= \frac{n_j}{n_{\jb}}n\bar{y}_j^2 \\
\end{align*}
and 
\begin{align*}
X_j^TX_k &= \sum_ix_{ij}x_{ik} \\
         &= \sum_i(p_{ij}\bar{y}_j + (1 - p_{ij})\bar{y}_{\jb})(p_{ik}\bar{y}_k + (1 - p_{ik})\bar{y}_{\kb}) \\
         &= \sum_i\left(p_{ij}p_{ik}\bar{y}_j\bar{y}_k + (1 - p_{ij})p_{ik}\bar{y}_{\jb}\bar{y}_k + p_{ij}(1 - p_{ik})\bar{y}_j\bar{y}_{\kb} + (1 - p_{ij})(1 - p_{ik})\bar{y}_{\jb}\bar{y}_{\kb}\right) \\
\end{align*}

These four terms represent the four outcomes in a two-by-two table, perhaps 
worth investigating further. Let:

  * $n_{jk} = \sum_ip_{ij}p_{ik}$,
  * $n_{{\jb}k} = \sum_i(1 - p_{ij})p_{ik}$,
  * $n_{j{\kb}} = \sum_ip_{ij}(1 - p_{ik})$,
  * $n_{{\jb}{\kb}} = \sum_i(1 - p_{ij})(1 - p_{ik})$,
  * $\bar{y}_{jk} = \frac{1}{n_{jk}}\sum_ip_{ij}p_{ik}y_i$,
  * $\bar{y}_{{\jb}k} = \frac{1}{n_{{\jb}k}}\sum_i(1 - p_{ij})p_{ik}y_i$,
  * $\bar{y}_{j{\kb}} = \frac{1}{n_{j{\kb}}}\sum_ip_{ij}(1 - p_{ik})y_i$,
  * $\bar{y}_{{\jb}{\kb}} = \frac{1}{n_{{\jb}{\kb}}}\sum_i(1 - p_{ij})(1 - p_{ik})y_i$.

These relate to the earlier quantities in the following ways

  * $n_j = n_{jk} + n_{j{\kb}}$, $n_{\jb} = n_{{\jb}k} + n_{{\jb}{\kb}}$,
  $n_{k} = n_{jk} + n_{{\jb}k}$, $n_{\kb} = n_{j{\kb}} + n_{{\jb}{\kb}}$,
  * $\bar{y}_j = \frac{n_{jk}}{n_j}\bar{y}_{jk} + \frac{n_{j{\kb}}}{n_j}\bar{y}_{j{\kb}}$, 
  $\bar{y}_{\jb} = \frac{n_{{\jb}k}}{n_{\jb}}\bar{y}_{{\jb}k} + \frac{n_{{\jb}{\kb}}}{n_{\jb}}\bar{y}_{{\jb}{\kb}}$, 
  $\bar{y}_k = \frac{n_{jk}}{n_k}\bar{y}_{jk} + \frac{n_{{\jb}k}}{n_k}\bar{y}_{{\jb}k}$, 
  $\bar{y}_{\kb} = \frac{n_{j{\kb}}}{n_{\kb}}\bar{y}_{j{\kb}} + \frac{n_{{\jb}{\kb}}}{n_{\kb}}\bar{y}_{{\jb}{\kb}}$.  

This two-way table notation allows us to write 

\begin{align*}
X_j^TX_k &= n_{jk}\bar{y}_j\bar{y}_k + n_{{\jb}k}\bar{y}_{\jb}\bar{y}_k + n_{j{\kb}}\bar{y}_j\bar{y}_{\kb} + n_{{\jb}{\kb}}\bar{y}_{\jb}\bar{y}_{\kb} \\
         &= n_{jk}\bar{y}_j\bar{y}_k + n_{{\jb}k}(-\frac{n_j}{n_{\jb}} \bar{y}_j)\bar{y}_k + n_{j{\kb}}\bar{y}_j(-\frac{n_k}{n_{\kb}} \bar{y}_k) + n_{{\jb}{\kb}}(-\frac{n_j}{n_{\jb}} \bar{y}_j)(-\frac{n_k}{n_{\kb}} \bar{y}_k) \\
         &= \bar{y}_j\bar{y}_k \left( n_{jk} - n_{{\jb}k}(\frac{n_j}{n_{\jb}}) - n_{j{\kb}}(\frac{n_k}{n_{\kb}}) + n_{{\jb}{\kb}}(\frac{n_j}{n_{\jb}})(\frac{n_k}{n_{\kb}}) \right) \\
\end{align*}

Notice that if $\frac{n_j}{n_{\jb}} = \frac{n_{jk}}{n_{{\jb}k}}$ then $X_j^TX = 0$.
This corresponds to the two genes being independant in presence/absence, which 
results in a covariance of zero.




Let $\eta$ be a $d$-dimensional unit vector i.e. with entries 
$\eta_j \in \mathbb{R}$ such that $\sum_j \eta_j^2 = 1$, and let $\mathbb{X}$ be 
the data matrix with entries $x_{ij}$, then consider 
\begin{align*}
(\eta \mathbb{X})^T \eta \mathbb{X} 
  &= \sum_i \left( \sum_j \eta_j x_{ij} \right)^2 \\
  &= \sum_i \sum_j \sum_k \eta_j \eta_k x_{ij} x_{ik} \\
  &= \sum_j \eta_j^2 \sum_i x_{ij}^2 + 2\sum_j \sum_{k < i} \eta_j \eta_k \sum_i x_{ij} x_{ik} \\
  &= \sum_j \eta_j^2 X_j^TX_j + 2\sum_j \sum_{k < i} \eta_j \eta_k X_j^TX_k \\
\end{align*}


## Read in and subset data

```{r}
# Read tidy data saved at the of initial_data_organisation.Rmd
load(file.path('data', 'pav_data.RData'))

# Reduce to NBS genes
nbs_table = pav_table[, c(TRUE, names(pav_table)[-1] %in% nbs$Name)]

# Merge on Yield and reduce to lines for which we have yield data
meta.df = subset(meta.df, !is.na(Yield))
nbs_table = merge(meta.df[, c('Line', 'Yield')], nbs_table)

# Reduce to genes with some decent level of variation
nbs_table = nbs_table[, c(TRUE, TRUE, colMeans(nbs_table[, -(1:2)]) <= 0.98 & colMeans(nbs_table[, -(1:2)]) >= 0.02)]

# Simplify gene names slightly
names(nbs_table) = sub('00.1.p$', '', names(nbs_table))
names(nbs_table) = sub('^GlymaLee.', 'GL', names(nbs_table))
names(nbs_table) = sub('^UWASoyPan', 'UWA', names(nbs_table))

# Move Line information into row.names
row.names(nbs_table) = nbs_table$Line
nbs_table$Line = NULL
```


