---
title: "NBS pairwise tests"
author: "Lyron Winderbaum"
date: "2021-03-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup}
# # Cleanup and Global Settings
# rm(list = ls())
# if (!is.null(sessionInfo()$otherPkgs)) {
#   invisible(lapply(paste0('package:', names(sessionInfo()$otherPkgs)),
#                    detach, character.only=TRUE, unload=TRUE))
# }
# graphics.off()
# options(stringsAsFactors = FALSE)

library(tidyverse)
```

## Read in and subset data

```{r}
# Read tidy data saved at the of initial_data_organisation.Rmd
load(file.path('data', 'pav_data.RData'))

# Reduce to NBS genes
nbs_table = pav_table[, c(TRUE, names(pav_table)[-1] %in% nbs$Name)]

# Merge on Yield and reduce to lines for which we have yield data
meta.df = subset(meta.df, !is.na(Yield))
nbs_table = merge(meta.df[, c('Line', 'Yield')], nbs_table)

# Reduce to genes with some decent level of variation
nbs_table = nbs_table[, c(TRUE, TRUE, colMeans(nbs_table[, -(1:2)]) <= 0.98 & colMeans(nbs_table[, -(1:2)]) >= 0.02)]

# # Proportion presence
# hist(colMeans(nbs_table[, -(1:2)]))


# Simplify gene names slightly
names(nbs_table) = sub('00.1.p$', '', names(nbs_table))
names(nbs_table) = sub('^GlymaLee.', 'GL', names(nbs_table))
names(nbs_table) = sub('^UWASoyPan', 'UWA', names(nbs_table))

# Move Line information into row.names
row.names(nbs_table) = nbs_table$Line
nbs_table$Line = NULL
```


# NBS Pairwise Tests

So, if we restrict to NBS genes that occur in at least 2% and no more than 98% 
of individuals in our dataset this leaves us with 76 genes (noting we shorten 
GlymaLee to GL and UWASoyPan to UWA for convenience):
```{r}
gene_nms = names(nbs_table)[-1]
gene_nms
```

We we consider all possible pairwise interactions (of which there are 2850), and 
perform a t-test for the interaction term in these linear models with Yield as 
the response:
```{r}
gene_nms = names(nbs_table)[-1]
df_pairs = data.frame()
for (i in 1:(length(gene_nms)-1)) {
  for (j in (i + 1):length(gene_nms)) {
    f = as.formula(paste('Yield ~', gene_nms[i], '*', gene_nms[j]))
    m = lm(f, data = nbs_table)
    int_nm = paste0(gene_nms[i], ':', gene_nms[j])
    if (int_nm %in% row.names(summary(m)$coefficients)) {
      df_pairs = rbind(
        df_pairs, 
        data.frame(gene.1 = gene_nms[i], gene.2 = gene_nms[j], 
                   p.val = summary(m)$coefficients[int_nm, "Pr(>|t|)"])
      )
    } else {
       df_pairs = rbind(
        df_pairs, 
        data.frame(gene.1 = gene_nms[i], gene.2 = gene_nms[j], p.val = NA)
      ) 
    }
  }
}

df_pairs = df_pairs[order(df_pairs$p.val),]
df_pairs = na.omit(df_pairs)
```

234 of these interactions cannot be computed (because there is no difference in 
gene presence/absence between the genes), leaving us with 2764 interactions.
Lets have a quick look at the p-value distribution:

```{r}
hist(df_pairs$p.val)
```

```{r}
sig_lvl = c(0.1, 0.05, 0.01)
rbind(sig_lvl,
      p_vals = sapply(sig_lvl, function(x){return(sum(df_pairs$p.val < x))}))
```

But we should apply a correction for multiple testing, using the Holm-Bonferroni 
method we get: 

```{r}
sig_lvl = c(0.1, 0.05, 0.01)
rbind(sig_lvl,
      p_vals = sapply(sig_lvl, function(x){return(sum(df_pairs$p.val*(nrow(df_pairs):1) < x))}))
```

So which are these three interesting interactions?

```{r}
subset(df_pairs, p.val*(nrow(df_pairs):1) < 0.05)
```



# Global Pairwise Tests

We can try the same procedure for all genes, not just the NBS genes. But first 
lets just tidy the full data in the same way we tidied the NBS data.

```{r}
pav_table = merge(meta.df[, c('Line', 'Yield')], pav_table)

# Reduce to genes with some decent level of variation
pav_table = pav_table[, c(TRUE, TRUE, colMeans(pav_table[, -(1:2)]) <= 0.98 & colMeans(pav_table[, -(1:2)]) >= 0.02)]
```

The filter to only include genes with proportion of occurance between 2% and 98%
reduces the 51415 genes down to just 2648 genes --- a huge number of these genes
are either very rare or very common, making it difficult to assess their 
imporance on Yield to to sample size limitations. Finish tidying:

```{r}
# Simplify gene names slightly
names(pav_table) = sub('00.1.p$', '', names(pav_table))
names(pav_table) = sub('^GlymaLee.', 'GL', names(pav_table))
names(pav_table) = sub('^UWASoyPan', 'UWA', names(pav_table))

# Move Line information into row.names
row.names(pav_table) = pav_table$Line
pav_table$Line = NULL
```

and we can repeat the pairwise test procedure:
```{r}
gene_nms = names(pav_table)[-1]
# source(file.path('analysis', 'pairwise_tests_all.R'))
df_pairs = read.csv(file.path('output', 'pairwise_tests_all.csv'))
df_pairs$gene.1 = gene_nms[df_pairs$gene.1]
df_pairs$gene.2 = gene_nms[df_pairs$gene.2]

df_pairs = df_pairs[order(df_pairs$p.val),]
df_pairs = na.omit(df_pairs)
```

Note I did some hackery there to run the code in a seperate script, store output, 
and load it in a particular format to get around re-running slow clode and get 
around the GitHub file size limit. 

Of the 3,499,335 pairs of genes, 373,817 cannot compute interaction terms because 
of the two genes having exact overlap. This leaves 3,125,518 pairs for testing.
Lets take a look at the p-value distribution:

```{r}
hist(df_pairs$p.val)
```

Nice, lookes like a clear addition between a uniform distribution and an 
exponential distribution, just what we would expect if there are some real 
interactions mixed in. Lets count the small p-values:

```{r}
sig_lvl = c(0.1, 0.05, 0.01)
rbind(sig_lvl,
      p_vals = sapply(sig_lvl, function(x){return(sum(df_pairs$p.val < x))}))
```

and adjust for multiple testing using the Holm-Bonferroni method: 

```{r}
sig_lvl = c(0.1, 0.05, 0.01)
rbind(sig_lvl,
      p_vals = sapply(sig_lvl, function(x){return(sum(df_pairs$p.val*(nrow(df_pairs):1) < x))}))
```

So which are these 38 interesting interactions significant at the 1% level after
multiple test correction?

```{r}
subset(df_pairs, p.val*(nrow(df_pairs):1) < 0.01)
```

Do any of these interactions involve NBS genes? Yes! UWA00005, GL15G1995, 
UWA01330, and UWA00202. Are any of these interactions between NBS genes? No! 
they are all pairs of one NBS gene to one non-NBS gene:

```{r}
subset(df_pairs, p.val*(nrow(df_pairs):1) < 0.01 & (gene.1 %in% names(nbs_table)[-1] | gene.2 %in% names(nbs_table)[-1]))
```


